---
title: "Data Analysis"
author: "Adejumo Ridwan Suleiman"
date: "`r Sys.Date()`"
output: pdf_document
---

## Importing and Exporting Data
- Before a data can be used in R, it needs to be imported and saved as a variable.
- A data is imported using the `read.csv()` function.
- The data will be working with is the sales data
```{r}
sales <- read.csv("C:/Users/abu aisha/Documents/R-Workshop-17-18-12-2022/sales.csv")
```
- The `read.csv()` takes the file path of the data as an argument.
- Data can also be exported using the `write.csv()` function.
- let's subset some columns in the data and export.
```{r}
subset <- sales[,c(2,3,4)]
write.csv(subset, "subset.csv")
```

# Analyzing Data 
- You can get a view of the whole data on a new tab using the `View()` function or use the`head()` function to get the first 5 rows.
```{r}
head(sales)
View(sales)
```

- You can check the data types of your data using the `str()` function.
```{r}
str(sales)
```
- You can check for the descriptive statistics using the `summary()` function
```{r}
summary(sales)
```
- Let's find out the total amount of all sales made.
```{r}
sum(sales["total"])
```
- Total payment by cash
```{r}
sum(sales[sales$payment == "Cash",]["total"])
```
- Most expensive item by unit price
```{r}
sales[sales$unit_price == 66.62,]
```
- Total sales made in each warehouse
```{r}
#Central
sum(sales[sales["warehouse"] == "Central",]["total"])
#North
sum(sales[sales["warehouse"] == "North",]["total"])
#West
sum(sales[sales["warehouse"] == "West",]["total"])
```

# Visualization
- Data visualization are performed in R using the
ggplot2 package.
- Let's find the frequency of sales made by warehouse.
```{r}
library(ggplot2)
qplot(x = warehouse,y = total, fill = warehouse, data = sales, geom = "col")
```
- Let's also find the total sales made by each client from the various warehouses.
```{r}
qplot(x = warehouse, 
      y = total, 
      fill = client_type, 
      data = sales, 
      geom = "col")
```
- Let's also see the payment system with the highest transactions.
```{r}
qplot(x = payment, 
      y = total, 
      fill = client_type, 
      data = sales, 
      geom = "col")
```



